---
title: "The anievent data structure"
format:
  html:
    toc: true
    toc-depth: 3
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(anicore)
```

## At a glance

An `anievent` is the sibling of [`anipoint`](aniframe-structure.html)
for **behavioural events**: sleep bouts, vocalisations, manual scoring
from BORIS, the output of segmentation models.

Each row is one **event**. An event sits in a **channel** (where it
is, as a soft convention, mutually exclusive with other events on the
same channel), has a **type** — either `"state"` (durative) or
`"point"` (instant) — and carries a **label** identifying which event
it is. Each event has a **start** and a **stop** time (identical for
point events). Just like an `anipoint`, the event is *performed by
something* (identity columns, declared via `variables_what`) at *some
time* (temporal-grouping columns, declared via `variables_when`).

Both classes inherit from the abstract `aniframe` parent, which carries
the shared substrate — so `get_metadata()`, `set_metadata()`, the dplyr
methods, and the rest of the family API work on both.

## Channels

A **channel** is a single mutually-exclusive categorical track of
behaviour. Within one channel, a subject can carry only one label at
a time: a sleep-state channel is either `REM` or `wake`, never both
simultaneously. If two behaviours **can** co-occur — say grooming
while walking — they belong on **separate channels**, not as two
values of the same channel. That mutual-exclusivity constraint is
what makes a channel a channel.

This means how you design your channels is a modelling choice. BORIS
calls them "behavioural categories" in the ethogram; segmentation
pipelines call them "label streams"; we call them channels in this
package.

## Anatomy at a glance

```{r}
ae <- anievent(
  individual = c(1L, 1L, 1L, 2L),
  channel = c("behaviour", "behaviour", "call", "behaviour"),
  label =c("REM", "wake", "alarm", "REM"),
  start = c(3, 14, 4.5, 1),
  stop = c(9, 19, 4.5, 6)
)
ae
```

Two things to spot:

1. The **header rows** prefixed with `#` come from the metadata
   attribute — identity columns and the event channels carried by the
   `channel` column.
2. The **data columns** are the long-format event records: one row per
   bout (or instant), with the channel name in `channel` and the
   label in `label`.

## Required schema

An `anievent` has five mandatory columns plus identity columns that
travel via `variables_what`:

| Column       | Type      | Required? | Role                                       |
|--------------|-----------|-----------|--------------------------------------------|
| `channel`    | character | yes       | which event channel                        |
| `type`       | factor    | yes       | `"state"` (durative) or `"point"` (instant) |
| `label`      | factor    | yes       | which event (the categorical label)        |
| `start`      | numeric   | yes       | bout start (or instant)                    |
| `stop`       | numeric   | yes       | bout stop (equals `start` for point events) |
| identity     | as source | optional  | e.g. `individual`, `track`, `subject`      |
| `modifiers`  | list-column of `character` | optional | flat per-event modifier values |

Identity columns are auto-detected from a known list
(`model`, `individual`, `subject`, `track`, `keypoint`) — the same list
`anipoint` uses, so identity means the same thing on both sides — or you
can supply any name via `variables_what`. An `anievent` with no identity
column at all is fine for single-subject experiments.

`type` is auto-derived from `start`/`stop` at construction.
Classification is **per `(channel, label)` group**, not per row: a
group is labelled `"point"` only when *all* of its bouts have
`start == stop`. If even one bout in the group is durative, the
whole group is `"state"` — so the same behaviour stays consistently
typed across its occurrences. Pass `type` explicitly to
override this default.

### Multiple observations / trials / sessions

Behavioural coding often spans several clips or trials, each with its
own time origin (`start = 0` means "the start of *this* clip", not "the
start of the experiment"). To carry that context, add an `observation`,
`session`, or `trial` column — those three names are auto-detected
as `when` keys, next to the `start`/`stop` interval, and the column is
coerced to factor / integer the same way anipoint does it:

```{r}
ae_multi <- anievent(
  individual = c(1L, 1L, 1L, 1L),
  observation = c("clip_a", "clip_a", "clip_b", "clip_b"),
  channel =c("behaviour", "behaviour", "behaviour", "behaviour"),
  label =c("REM", "wake", "REM", "wake"),
  start = c(3, 14, 1, 7),
  stop = c(9, 19, 5, 12)
)
get_variables_when(ae_multi)
ae_multi
```

The `start`/`stop` values are local to each observation — both clips
have an event starting near the beginning, but those are separate
moments in real time. Anything downstream that needs to combine clips
onto a shared timeline (plots, cross-clip metrics) will need an
observation-level offset table; the class itself just carries the key.

### Time unit

`unit_time` is carried in the shared metadata substrate, so an
`anievent` declares the unit of its `start`/`stop` values just like
an `anipoint` declares the unit of its
[index column](aniframe-structure.html#the-index) (`"frame"`,
`"ms"`, `"s"`, `"m"`, `"h"`, …). `set_unit_time()` and
`set_sampling_rate()` dispatch on both classes — on an anievent they
scale `start` and `stop` instead of the index:

```{r}
ae_frames <- anievent(
  individual = 1L,
  channel = c("behaviour", "behaviour"),
  label =c("REM", "wake"),
  start = c(30, 150),
  stop = c(60, 300)
)
ae_seconds <- set_sampling_rate(ae_frames, sampling_rate = 30)
ae_seconds
```

## State vs point events

Two temporal kinds:

- **State events** — interval-valued, durative (`stop > start`). A
  sleep bout, a grooming bout.
- **Point events** — instantaneous (`stop == start`). A startle, a
  vocalisation onset.

The `anievent` class is intentionally **type-agnostic**: it does not
require you to declare which channels are state and which are point.
The `channel` column already names the channels, and the duration of
each row tells you everything the class needs.

Type only becomes load-bearing at conversion / plotting / metric time:

- when converting an `anievent` to a per-frame `anipoint` (so the
  conversion knows whether to fill the whole interval or just a single
  frame),
- when plotting (state → coloured band; point → marker),
- when computing metrics that differ by type.

Those operations live in companion packages (`animetric`, `anivis`)
and will accept type declarations at call time. On the `anipoint`
side the distinction can be recorded explicitly with
`set_variables_event(state = ..., point = ...)`, but the `anievent`
class itself does not enforce it — an anievent has no
`variables_event` of its own, since its events are already encoded in
`channel` and `label`.

## Modifiers

A **modifier** is a per-event attribute set at coding time — in BORIS,
the values chosen from a behaviour's modifier sets. A grooming bout
might carry the values `"limb"` and `"whisker"`; an alarm call might
carry `"high"`.

Modifiers are stored as an optional `modifiers` list-column, where
each cell is a **flat character vector** (length zero, one, or many).
This matches how BORIS exports modifier data: the set-to-value
structure from the ethogram is collapsed into a single bag of values
per event, so a modifier cell is just the values picked, with no
set-key wrapping.

```{r}
ae_with_mods <- anievent(
  individual = c(1L, 1L, 1L),
  channel =c("behaviour", "behaviour", "call"),
  label =c("REM", "REM", "alarm"),
  start = c(3, 14, 4.5),
  stop = c(9, 19, 4.5),
  modifiers = list(
    c("limb", "whisker"),
    "tail",
    character()
  )
)
ae_with_mods
```

Standard dplyr verbs round-trip the list-column without further
ceremony:

```{r}
ae_with_mods |>
  dplyr::filter(channel == "behaviour")
```

A helper to widen this list-column into one logical column per
distinct modifier value (so filtering or faceting on a single
modifier value is straightforward) is planned but not yet
implemented.

## Validation

`validate_anievent()` re-checks the structural invariants on demand,
returning the input invisibly on success and erroring otherwise:

```{r}
#| error: true
# Happy path: returns invisibly, no output
validate_anievent(ae_with_mods)

# Negative interval is caught
bad <- ae_with_mods
bad$stop[1] <- bad$start[1] - 1
validate_anievent(bad)
```

The validator confirms required columns are present with the right
types, that `stop >= start` for every row, and that any `modifiers`
column is well-formed. It deliberately does **not** check
type-linked invariants (e.g. point rows having `start == stop`) —
those belong to the conversion code that knows which channels are
typed as what.

## Encoding per-frame data into an `anievent`

Per-frame data (factor / character / logical columns over a time
grid) is encoded into the bout shape with `to_anievent()`. The verb
is distinct from `as_anievent()`: that one is a *strict cast* of an
already-bout-shaped data frame (with `channel` / `type` / `label` /
`start` / `stop`); `to_anievent()` *produces* that shape via
run-length encoding.

### From a data frame

`to_anievent()` on a data frame takes bare-name selections for the
event columns and the time / identity context.

```{r}
library(tibble)
df <- tibble(
  individual = 1L,
  time       = 1:8,
  behaviour  = factor(c("REM", "REM", "REM", "wake", "wake", "REM", "REM", NA)),
  woke_up    = c(FALSE, FALSE, FALSE, TRUE, FALSE, FALSE, FALSE, FALSE),
  call       = c(NA, "alarm", NA, NA, NA, NA, NA, NA)
)

to_anievent(
  df,
  time           = time,
  state          = c(behaviour, woke_up),
  point          = call,
  variables_what = individual
)
```

Three things to notice:

- `behaviour` is a multi-level factor → one channel with both `REM`
  and `wake` labels, RLE'd into contiguous bouts.
- `woke_up` is a logical → TRUE-runs become bouts labelled
  `"woke_up"` (the column name); FALSE frames produce nothing.
- `call` is a sparse factor passed to `point` → one bout per
  non-`NA` frame with `start == stop`.

The channel name always comes from the column name. Use `state` vs
`point` to pick how the column is interpreted — the value-shape
(factor / character / logical) is handled internally.

### From an `anipoint`

If you already have an `anipoint` with `variables_event` declared in
metadata, `to_anievent()` reads everything from there:

```{r}
af <- anipoint(
  individual = rep(1L, 8),
  time = 1:8,
  x = 1:8,
  y = 1:8,
  behaviour = factor(
    c("REM", "REM", "REM", "wake", "wake", "REM", "REM", NA),
    levels = c("REM", "wake")
  )
)
af <- set_variables_event(af, state = "behaviour")
to_anievent(af)
```

The resulting anievent inherits `unit_time` and `sampling_rate` from
the host. `stop` is the time of the last frame of each bout
(inclusive, frame-aligned). If a `<col>_modifiers` list-column is
present alongside an event column, its cells are gathered into the
anievent's `modifiers` column, one entry per emitted bout.

The reverse direction (`anievent` → `anipoint`) is intentionally not
provided yet — the anievent's loose mutual-exclusion semantics don't
map onto an anipoint's per-frame columns without a design decision
about how to resolve overlap, and that design is still pending.

## Where things live in the ecosystem

| Concern                                 | Planned home | Status     |
|-----------------------------------------|--------------|------------|
| The `anievent` class                    | `anipoint`   | ✅ implemented |
| Encoding per-frame data (`to_anievent()`) | `anipoint`   | ✅ implemented |
| Conversion to `anipoint`                | `anipoint`   | planned    |
| Readers (BORIS, Solomon, …)             | `aniread`    | planned    |
| Bout-summary metrics                    | `animetric`  | planned    |
| Plots (state band, point marker)        | `anivis`     | planned    | -->
