---
title: "Metadata on an aniframe"
format:
  html:
    toc: true
    toc-depth: 3
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(anicore)
```

## Why an attribute, not columns?

Every `anipoint` carries a metadata list as an R attribute alongside
the data columns. The metadata records the things that are true of the
recording as a whole rather than of any single observation: the source
software, the sampling rate, what units the spatial coordinates are
in, which way its axes point, and so on.

Keeping this information attached to the object — rather than living
in a separate file or being passed around as extra arguments — is what
lets the rest of the _animovement_ ecosystem stay loosely coupled.
A reader (in _aniread_) populates the metadata at load time, and any
downstream tool can read it back without a hand-off.

This article covers the metadata attribute and the functions that read
and update it. For the data-column structure see
[The aniframe data structure](aniframe-structure.html); for
connections, which live in the `structure` category, see
[Connections](aniframe-connections.html).

## The metadata attribute

You can see the full metadata by printing it directly:

```{r}
data <- example_anipoint()
get_metadata(data)
```

The fields and their defaults are defined in one place,
`list_default_metadata()` — that's the canonical source of truth for what
an `anipoint`'s metadata looks like.

```{r}
str(list_default_metadata(), max.level = 2)
```

The fields are grouped into **categories**, each answering one kind of
question:

| Category      | Fields                                                              |
|---------------|---------------------------------------------------------------------|
| `recording`   | provenance: `source`, `source_version`, `source_format`, `filename` |
| `time`        | `unit_time`, `sampling_rate`, `sampling_interval`, `start_datetime` |
| `space`       | `coordinate_system`, `reference_frame`, `handedness`, `axis_directions`, `axis_extents`, `unit_space`, `unit_angle` |
| `variables`   | which columns play which role: `what`, `when`, `where`, `event`, each a list of named slots |
| `structure`   | relationships between levels of a variable — today's connections    |

`spec_version` sits above the categories: it is metadata about the
metadata, versioning the contract they describe. `space` is the one
category a class can lack — an `anievent` has no spatial component, so
its metadata simply has no `space` rather than a set of "none" values.

`filename` accepts a character vector — readers like
`aniread::read_trackball()` populate it with all source paths.

## Reading and writing metadata

`get_metadata()` and `set_metadata()` are the workhorses. The categories
group the storage, but access stays **flat**: a field is found and
written by its own name wherever it lives, and a category name returns
the whole category.

```{r}
get_metadata(data, "sampling_rate")

data <- set_metadata(data, sampling_rate = 30, source = "deeplabcut")
get_metadata(data, "sampling_rate")
get_metadata(data, "source")

names(get_metadata(data, "space"))
```

`set_metadata()` validates the input — factor fields are checked
against their permitted levels, and unknown fields are rejected.

For fields whose update has side effects on the data columns (or on
related fields), prefer the dedicated setters listed below.

| Setter                | Touches                                                        |
|-----------------------|----------------------------------------------------------------|
| `set_unit_space()`    | converts `x`/`y`/`z` between length units                      |
| `set_unit_time()`     | converts the index column between time units                   |
| `set_unit_angle()`    | converts `phi`/`theta` (auto) and any extra `cols` you supply  |
| `set_sampling_rate()` | flips `unit_time` from frames to seconds and rescales the index |
| `set_index()`         | changes which column the frame is indexed by, and re-orders it |
| `set_axes()`          | declares which column carries which axis role                  |
| `set_axis_directions()` | says which way each axis points, reflecting one turned over |
| `set_axis_extents()`  | says how far each axis runs, which a reflection turns around   |
| `set_handedness()`    | says whether the frame is right- or left-handed                |

The temporal setters follow the frame's own declaration rather than a
column called `time` — see
[the index](aniframe-structure.html#the-index).

## Declaring the slot vocabulary

The `variables` category is a special case: its roles name columns
rather than describing values, so a name that matches nothing is a
promise the frame can't keep. All but the `event` role go further — they
are not a description of the frame, they *are* its structure.
`as_anipoint()` uses them to coerce column types, order columns and
rows, group the frame, and derive `coordinate_system`. Writing them
without redoing that work would leave the frame and its own metadata
disagreeing — the print header would update while the grouping still
reflected the old declaration.

The category holds one list per role, each with named slots:
`what$keys` (identity), `when$index` and `when$keys` (the position
within a context, and the context itself — an `anievent` has
`when$interval` instead of an index), `where$position` (the axis-role
mapping), and `event$state` / `event$point`. The frame groups by
`c(what$keys, when$keys)` and nothing else.

`set_metadata()` refuses the whole category, and points you at the
setters that do the whole job:

```{r error=TRUE}
data |> set_metadata(variables_what = "id")
```

The three structural roles have the same four verbs as `connections`:
`get_variables_*()`, `set_variables_*()`, `add_variables_*()` and
`remove_variables_*()`. `variables_index` has only two — `get_index()`
and `set_index()` — because a frame has exactly one index, leaving
nothing for `add_` and `remove_` to do. The column has to exist before
it can be declared, so the order is always create-then-declare:

```{r}
tagged <- data |>
  dplyr::mutate(id = "trial_1") |>
  add_variables_what("id")

get_variables_what(tagged)
dplyr::group_vars(tagged)
```

`add_variables_*()` appends to the declaration, so you don't have to
restate what is already there — forgetting to would quietly demote an
existing identity variable and regroup the frame without it.

Declaring a spatial column refreshes the fields derived from it:

```{r}
data |>
  dplyr::mutate(z = 0) |>
  add_variables_where("z") |>
  get_metadata("coordinate_system")
```

`variables_event` is the fourth role, and the one that doesn't change
the frame's shape: it declares which columns carry per-frame event
labels, split into interval-valued `state` columns and instantaneous
`point` columns. `to_anievent()` reads it to know what to encode.

```{r}
data |>
  dplyr::mutate(behaviour = factor("rest")) |>
  set_variables_event(state = "behaviour") |>
  get_variables_event()
```

## Which way the axes point

For 2D image-derived data there's an annoying convention split: most
image and video tooling counts y **downward** from the top of the frame,
while plotting and most maths counts it **upward**. What differs is not
where `(0, 0)` sits — it is a corner either way — but the direction y
increases in, so that is what `anipoint` records.

`axis_directions` maps each axis role to one of six words, in three
opposed pairs, read from where the recording was made: `right`/`left`
across the view, `up`/`down` within it, and `back`/`forward` toward and
away from the viewer.

Turning an axis to its opposite reflects that column, so the data ends up
expressed the way you just declared. An axis runs from zero to its
`axis_extents` value, so the reflection is `extent - old` — and an axis
with no declared extent is centred on its origin instead, so turning it
over negates it.

```{r}
img <- anipoint(
  individual = 1L, time = 1:4,
  x = c(0, 10, 20, 30),
  y = c(50, 100, 150, 200)
) |>
  set_axis_extents(c(y = 1080)) |>
  set_axis_directions(c(x = "right", y = "down"))

img$y
img <- set_axis_directions(img, c(y = "up"))
img$y # reflected: 1080 - original y
```

### What follows from the directions

Two things are read off them rather than recorded separately, so they
cannot go on claiming a convention the axes no longer have.

`get_angle_direction()` says which way angles run. `atan2(y, x)` counts
counter-clockwise, so the same physical heading comes out mirrored
between a y-down and a y-up frame — which is exactly the comparison that
goes wrong silently without this.

`get_handedness()` needs three axes. Two directions leave it open, so a
frame with only `x` and `y` has none until it says which side it was
observed from — and that is not a detail. A rodent filmed from above and
the same rodent filmed from below through a glass floor give images whose
`x` and `y` are declared identically, but whose rotations run opposite
ways. The depth axis is the only thing that tells them apart.

```{r}
above <- set_axis_directions(img, c(z = "back"))
below <- set_axis_directions(img, c(z = "forward"))

c(above = get_angle_direction(above), below = get_angle_direction(below))
c(above = get_handedness(above), below = get_handedness(below))
```

Since `det[x y z]` is `(x × y) · z`, a right-handed frame counts
counter-clockwise about its own depth axis — always. The two answers are
one fact seen twice.

A frame can also state the convention without spelling the axes out,
which is how most 3D recordings are described:

```{r}
set_handedness(img) |> get_handedness() # right-handed by default
```

## Units

Spatial, temporal, and angular units each have their own setter.
Conversions between standard units are automatic; conversions from
unknown / `frame` / `px` units require an explicit calibration factor
(or, for time, a `sampling_rate`).

```{r}
data <- example_anipoint(n_dims = 2) |>
  set_metadata(unit_space = "mm")

data_cm <- set_unit_space(data, to_unit = "cm")
get_metadata(data_cm, "unit_space")
```

```{r}
data <- example_anipoint() # default unit_time = "frame"
data_s <- set_sampling_rate(data, sampling_rate = 30)
get_metadata(data_s, "unit_time") # now "s"
range(data_s$time) # frames divided by fps
```

Spatial angular columns (`phi`, `theta`) are converted automatically
by `set_unit_angle()` whenever they're present. Pass `cols` only for
non-spatial angular columns (e.g. heading direction).

```{r}
pol <- anipoint(
  individual = 1L, time = 1:3,
  rho = c(1, 1, 1), phi = c(0, pi / 2, pi)
)
pol$phi
pol_deg <- set_unit_angle(pol, to_unit = "deg")
pol_deg$phi
```
