---
title: "Connections"
format:
  html:
    toc: true
    toc-depth: 3
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(anicore)
```

## What `connections` are for

Many movement datasets need to record relationships *between* the
values of an identity or temporal variable. The most common case is a
**skeleton** — edges between keypoints — but the same shape works for
a social network (edges between individuals) or a session graph.

`anipoint` stores this in the `structure` category of its metadata,
managed by a small family of functions. This article covers those.

For the rest of the metadata see
[Metadata on an aniframe](aniframe-metadata.html); for the data-column
structure see [The aniframe data structure](aniframe-structure.html).

## Storage shape

`connections` is a **named list keyed by the variable** the
relationships describe. Each entry is a 2-column `from`/`to` tibble.
The order of `from` / `to` is preserved as supplied, so downstream
consumers can treat the table as either directed (parent → child
kinematic chains) or undirected (skeleton edges).

```{r}
fish <- example_anipoint(n_keypoints = 7) |>
  set_connections(
    list(
      c("head", "neck"),
      c("neck", "shoulder_right"),
      c("neck", "shoulder_left"),
      c("shoulder_right", "abdomen"),
      c("shoulder_left", "abdomen")
    )
  )

get_connections(fish, "keypoint")
```

## The four functions

```{r}
#| eval: false
set_connections(data, connections, variable = "keypoint") # replace
get_connections(data, variable = NULL) # full list, or one entry
add_connections(data, from, to, variable = "keypoint") # append (vectorised)
remove_connections(data, from, to, variable = "keypoint") # exact match
```

### `set_connections()`

Replaces the connection table for a single variable. Three input
shapes are accepted — pick whichever is most natural:

```{r}
#| eval: false
# Implicit by position: element[1] is `from`, element[2] is `to`
set_connections(data, list(c("head", "neck"), c("neck", "shoulder_right")))

# Named within each pair — use when direction should be obvious in the call
set_connections(data, list(c(from = "head", to = "neck")))

# 2-column data.frame — useful when connections come from another tabular source
set_connections(
  data,
  data.frame(
    from = c("head", "neck"),
    to   = c("neck", "shoulder_right")
  )
)
```

### `get_connections()`

Returns the full named list when `variable` is `NULL`, or a single
tibble when a variable name is given.

```{r}
get_connections(fish, "keypoint")
```

### `add_connections()` / `remove_connections()`

These mirror each other: both take `from` and `to` as either single
strings or vectors of equal length, so multiple edges go in or out in
one call.

```{r}
fish <- add_connections(
  fish,
  from = c("abdomen", "abdomen"),
  to   = c("hip_right", "hip_left")
)
nrow(get_connections(fish, "keypoint"))

# Same shape, in reverse — remove three pairs at once:
fish <- remove_connections(
  fish,
  from = c("abdomen", "abdomen", "head"),
  to   = c("hip_right", "hip_left", "neck")
)
nrow(get_connections(fish, "keypoint"))
```

`remove_connections()` matches exactly on `from`/`to`. Direction
matters: removing `(a, b)` won't strip `(b, a)`. If you stored an edge
in one direction but want it gone regardless of orientation, include
both pairs in the call (or invoke twice with swapped arguments).

## Connections on other variables

`connections` is keyed by *any* identity or temporal variable, not just
keypoints. A social-network study, for example, might record edges
between individuals:

```{r}
social <- example_anipoint(n_individuals = 3) |>
  set_connections(
    list(c("1", "2"), c("2", "3")),
    variable = "individual"
  )

get_connections(social)
```

The `variable` argument must be a column declared as a `what` or
`when` key — `set_connections()` errors otherwise to catch
typos.

## Catching typos

When a `from`/`to` value isn't found in the corresponding column,
`set_connections()` and `add_connections()` emit a warning but **keep
the connection** — the value may legitimately be missing in this
particular recording while being valid elsewhere.

```{r}
#| warning: true
example_anipoint(n_keypoints = 5) |>
  set_connections(list(c("head", "necc"))) # typo
```
