---
title: "Orientation"
format:
  html:
    toc: true
    toc-depth: 3
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(anicore)
```

Orientation means two things on an `anipoint`, and this article covers
both. The first is the frame's own: which way its axes point, how far
they run, and what follows from that for angles and handedness. The
second is the entity's: which way an animal faces. The second is
measured against the first, so it comes after.

For the metadata fields in general see
[Metadata on an aniframe](aniframe-metadata.html); for position and the
`where` role see [The anipoint data structure](aniframe-structure.html).

## Which way the axes point

Image and video tooling counts `y` **downward** from the top of the
frame; plotting and most maths count it **upward**. The two disagree
about a direction, not about where `(0, 0)` sits — that is a corner
either way — so a direction is what `anipoint` records. A corner also
says nothing about a third axis, and a direction does.

`axis_directions` maps each axis role to one of six words, in three
opposed pairs, read from where the recording was made:

| Pair       | Directions        | Meaning                              |
|------------|-------------------|--------------------------------------|
| horizontal | `right` / `left`  | across the view                      |
| vertical   | `up` / `down`     | within the view                      |
| depth      | `back` / `forward`| toward / away from the viewer        |

The words are relative to the viewpoint, not to the world. Whether the
view is fixed in the arena or rides on the animal is `reference_frame`'s
job (`allocentric` or `egocentric`).

`set_axis_directions()` declares them, keyed by role. Roles not named
keep what they had, and `NA` clears one. It only declares: the values
are left alone.

```{r}
img <- anipoint(
  individual = 1L, time = 1:3,
  x = c(100, 200, 300),
  y = c(100, 100, 200)
) |>
  set_axis_directions(c(x = "right", y = "down"))

get_axis_directions(img)
img$y
```

Two axes along the same pair would be parallel, so that is refused:

```{r error=TRUE}
set_axis_directions(img, c(y = "left"))
```

### Roles, directions and extents

Three fields are keyed the same way and say different things. The `where`
role says which column carries `y` (`get_axes()`); `axis_directions`
says which way `y` points; `axis_extents` says how far it runs.

```{r}
img <- set_metadata(img, axis_extents = c(y = 1080))

get_axes(img)
get_axis_directions(img)
get_metadata(img, "axis_extents")
```

An extent is the length of the axis from zero — for image data, the
frame's height or width in pixels. Its one use is as the thing an axis
is reflected around. `set_metadata()` warns when the data already runs
past it, since that usually means the extent belongs to another
recording.

## Turning an axis over

`reflect_axis()` changes the data. It reflects the column carrying the
role and flips its declared direction, so the frame describes the same
scene with that axis pointing the other way. Around a declared extent
the reflection is `extent - old`:

```{r}
up <- reflect_axis(img, "y")

up$y
get_axis_directions(up)
```

Without an extent, the axis is centred on its origin and reflecting it
negates it:

```{r}
img |>
  set_metadata(axis_extents = c(y = NA)) |>
  reflect_axis("y") |>
  dplyr::pull(y)
```

Turning `y` from down to up is the common case, and aniread's readers do
it on import for image-derived formats. A known handedness flips too,
and so does any declared entity orientation (see
[below](#which-way-an-entity-faces)).

## What follows from the directions

Two things are read off the directions rather than stored, so they
cannot go on claiming a convention the axes no longer have.

`get_angle_direction()` is the sense of rotation from `x` toward `y`.
`atan2(y, x)` counts counter-clockwise, so the same physical heading
comes out mirrored between a y-down and a y-up frame:

```{r}
c(image = get_angle_direction(img), flipped = get_angle_direction(up))
```

`get_handedness()` needs three axes. Two directions leave it open, so
a 2D frame has a sense of rotation but no handedness:

```{r}
get_handedness(up)
```

That gap matters more than it looks. A rodent filmed from above and the
same rodent filmed from below through a glass floor give images whose
`x` and `y` are declared identically, but whose rotations run opposite
ways: a left turn is a rising angle in one and a falling angle in the
other. The depth axis is the only thing that tells them apart. With `z`
pointing up out of the floor, it points toward the camera above
(`back`) and away from the camera below (`forward`):

```{r}
above <- set_axis_directions(up, c(z = "back"))
below <- set_axis_directions(up, c(z = "forward"))

c(above = get_angle_direction(above), below = get_angle_direction(below))
c(above = get_handedness(above), below = get_handedness(below))
```

The angle direction is seen from the side `z` points to. Since
`det[x y z]` is `(x × y) · z`, a right-handed frame counts
counter-clockwise about its own depth axis — always. The two answers
are one fact seen twice.

Most 3D recordings state the convention rather than spelling out each
axis. Declare it directly; three declared directions take precedence
over it when both are present:

```{r}
right <- set_metadata(up, handedness = "right")
get_handedness(right)
get_angle_direction(right)
```

## Which way an entity faces

Position says where an entity is; orientation says which way it faces.
It is declared next to position, in the `where` role's `orientation`
slot, with the same pattern: a closed set of roles mapped to columns of
any name.

A 2D frame uses `yaw`, measured from `x` toward `y` in `unit_angle`:

```{r}
fly <- anipoint(
  individual = 1L, time = 1:3,
  x = c(0, 1, 2), y = c(0, 0, 1),
  heading = c(0, pi / 4, pi / 2)
) |>
  set_variables(where = list(orientation = c(yaw = "heading")))

get_variables(fly, "where", "orientation")
```

A 3D frame uses a unit quaternion, `qw`, `qx`, `qy`, `qz` (Hamilton
convention, scalar first). This one turns a quarter-turn about `z`
between the two rows:

```{r}
body <- anipoint(
  individual = 1L, time = 1:2,
  x = 0, y = 0, z = 0,
  qw = c(1, cos(pi / 4)), qx = 0, qy = 0, qz = c(0, sin(pi / 4))
) |>
  set_variables(
    where = list(orientation = c(qw = "qw", qx = "qx", qy = "qy", qz = "qz"))
  )

get_variables(body, "where", "orientation")
```

The declaration is checked against the frame. A quaternion needs a 3D
frame and `yaw` a 2D one, and a quaternion must have unit norm:

```{r error=TRUE}
body |>
  dplyr::mutate(qx = 0.5) |>
  set_variables(
    where = list(orientation = c(qw = "qw", qx = "qx", qy = "qy", qz = "qz"))
  )
```

Declared orientation follows the frame. `reflect_axis()` reflects it with
the positions — `yaw` like an angle, a quaternion by negating the two
components off the reflected axis — and `convert_unit_angle()` converts
`yaw`:

```{r}
reflect_axis(fly, "y")$heading
convert_unit_angle(fly, "deg")$heading
reflect_axis(body, "x")$qz
```

### Why quaternions, not Euler angles

Euler angles are how most sources report 3D orientation, and they break
under ordinary operations. Each angle wraps at ±π, so a mean or an
interpolation across the wrap gives the opposite direction:

```{r}
roll <- c(3.1, -3.1)
mean(roll)
circ_mean(roll)
```

`circ_mean()` fixes one column, but three Euler angles are not three
independent circular variables: near gimbal lock two of them describe
the same rotation, and no per-column treatment recovers it. A
`summarise()` or an interpolation over Euler columns gives a wrong
rotation without complaint. A quaternion has neither problem, so that is
what the frame stores.

Showing and entering orientation is a different matter, and there the
source's convention is what people know. A frame can record it, so tools
don't need it retyped each time:

```{r}
body <- set_metadata(body, euler_sequence = "ZYX", euler_intrinsic = TRUE)
get_metadata(body, "euler_sequence")
```

The two are set together. anispace's `transform_quaternion_to_euler()`
and `transform_euler_to_quaternion()` read them.

## Heading is not direction of travel

Orientation records measured state: FicTrac's heading (integrated from
the ball's rotation), a TRex posture angle, a rigid body from motion
capture. The direction of travel
is a different quantity, derived from successive positions, and the two
disagree exactly when it is interesting — side-slip, backing up, drift.
Keep derived travel direction in an ordinary column, not in
`where$orientation`.

Nor is every angle a tracker reports a `yaw`. Blob and mask trackers
such as Bonsai and Octron give the angle of an ellipse's long axis,
which has no front: θ and θ + π are the same. Declared as `yaw`, a
mean, a difference or a reflection would mix front and back, so these
stay as undeclared columns for now (see
[#165](https://github.com/animovement/anicore/issues/165)).
