---
title: "Structures"
format:
  html:
    toc: true
    toc-depth: 3
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(anicore)
```

## What a structure is

An `anistructure` records how the levels of one variable relate. The
common case is a **skeleton** over `keypoint`, but the same class holds
a team, a colony or a social network over `individual`. It has three
parts:

- **points**: the members.
- **segments**: directed links between two points, with an optional
  expected `length`.
- **joints**: ordered pairs of segments, with angular limits per degree
  of freedom.

Each part beyond the points is optional. A structure with points alone
is a roster; with segments it is a graph; with joints it is a
constrained skeleton.

```{r}
leg <- anistructure(
  segments = data.frame(
    segment = c("thigh", "shin"),
    from = c("hip", "knee"),
    to = c("knee", "ankle")
  ),
  joints = data.frame(joint = "knee", a = "thigh", b = "shin", min = 0, max = 2.5),
  root = "hip"
)
leg
```

Segments point one way. Run limbs from the root outward, and paired
segments (left to right) the same way, so joint angles come out with the
clinical convention: 0 when two segments are aligned.

## Attaching structures to a frame

A structure is a template: it names no variable until it is attached.
`set_structure()` binds it to a variable and stores it under a name,
which defaults to the variable:

```{r}
af <- example_anipoint(n_individuals = 4) |>
  set_structure(example_structure())

get_structure(af, "keypoint")
```

## Several structures over one variable

A frame can hold any number of named structures, and several may span
the same variable. A structure's points are its members, so a subset of
the levels makes a sub-group, and groups may overlap:

```{r}
af <- af |>
  set_structure(
    anistructure(points = c("1", "2", "3", "4")),
    variable = "individual",
    name = "team"
  ) |>
  set_structure(
    anistructure(segments = list(c("1", "2"), c("2", "3"))),
    variable = "individual",
    name = "defence"
  )

names(get_structure(af))
get_structure(af, "defence")
```

`remove_structure()` drops one by name.

Membership is fixed for the frame: a player who changes role partway
through needs a separate frame, or a column recording the role.

## From points to segments

`as_anisegment()` re-expresses an anipoint as the segments of one of its
structures. Each row is one segment at one time: its `length`, and the
unit vector from its `from` point to its `to` point (`ux`, `uy`, and
`uz` in 3D). The structure's variable is replaced by a `segment` key.

```{r}
skeleton <- example_anipoint(n_obs = 3, n_individuals = 1) |>
  set_structure(example_structure())

seg <- as_anisegment(skeleton)
seg
get_variables(seg, "where")
```

Segment lengths should stay constant for a rigid body, so their spread is
a first check on tracking quality:

```{r}
seg |>
  dplyr::summarise(length_sd = sd(length), .groups = "drop") |>
  head()
```

Editing the segments and rebuilding the positions carries the edit into
every point but the root. Holding each segment at its median length makes
the body rigid:

```{r}
rigid <- seg |>
  dplyr::mutate(length = stats::median(length)) |>
  as_anipoint(root = skeleton)
rigid
```

The rebuild walks the segments outward from the structure's `root`, so it
needs the root point's trajectory; `root` can be the frame the segments
came from.

## Catching typos

Points that don't appear in the data are kept with a warning, since
they may be recorded in another file:

```{r}
#| warning: true
example_anipoint(n_keypoints = 5) |>
  set_structure(anistructure(segments = list(c("head", "necc"))))
```
